Manuscript title here

### 2024-09-29 project-plan.Rmd
fixat de flesta källor, två kvar och kanske lägga några extra om jag hinner i trauma delen

Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate]. Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa). In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old. (Swetrau ÅR, 2021). This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs.

@@

Trauma are generally divided into two groups, blunt and penetrating injury. The most common type in Sweden is blunt injury, constituting around 90% of all trauma. The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma(Swetrau ÅR, 2021).

Blunt and penetrating trauma have their own distinct clinical implications. Blunt trauma often results from impacts of different energy factor, and depending on the energy can result in anything from abbrations to fractures and internal organ damage. The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), at 75% of all blunt trauma[@isenhour_advances_2007]. The most common cause is traffic related such as, rapid deceleration leading to driver crashing into car furniture and causing contusions or rupture of internal organs.

Penetrating trauma on the other hand are often caused by gunshots or stab wounds[@internet_archive_emergency_2002]. The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and heart tamponade.

@@

The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. Some of the common symptoms are tachycardia, hypotenstion and altered mental status. Shock are usually classified into four different types based on physiological mechanism. The types are, hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics(källa). These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system. Common examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - pulmonary embolism, cardiogenic - STEMI  and, distributive - sepsis(källa).

The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage[@teixeira_preventable_2007]. Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities(källa).

@@

To differentiate the severity of trauma shock, we can divide it into different degrees. One of such classification is the ATLS classification for major haemorrhage. ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).

@@

The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24 hour of injury[@berry_shock_2015], as the midian time to death is 2 hours[@tisherman_detailed_2015]. To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated. QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009]. By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally[@juillard_establishing_2009][@hashmi_hospital-based_2013].

@@

To classify, all of the parameters in one class needs to be fulfilled. In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017]. At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[ATLS].

>>>>

Trauma is a major global public health concern. It causes over four million death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate]. Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger demographic are also affected. In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old[@noauthor_arsrapport_2022]. This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs[@noauthor_injuries_nodate].

@@

Trauma are generally divided into two groups, blunt and penetrating injury. The most common type in Sweden is blunt injury, constituting around 90% of all trauma. The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma[@noauthor_arsrapport_2022].

Blunt and penetrating trauma have their own distinct clinical implications. Blunt trauma often results from impacts of different energy factors, and depending on the energy can result in anything from abbrations to fractures and internal organ damage. The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), which is often caused by traffic accidents. When counting motor vehicle collision and pedestrian versus auto accidents together, they account for 75% of all blunt trauma[@isenhour_advances_2007].

Penetrating trauma on the other hand are often caused by gunshots or stab wounds[@internet_archive_emergency_2002]. The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and cardiac tamponade.

@@

The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. Some of the common symptoms are tachycardia, hypotenstion and altered mental status. Shock are usually classified into four different types based on physiological mechanism. The types are: hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics[@standl_nomenclature_2018]. These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system. Examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - tension pneumothorax, cardiogenic - myocardial pump failure and distributive - sepsis[@standl_nomenclature_2018].

The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage[@teixeira_preventable_2007]. Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities[@raja_initial_2024].

@@

To differentiate the severity of trauma shock, we can divide it into different degrees. One of such classification is the ATLS classification for major haemorrhage. ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output[@noauthor_atls_2018]. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings, still is one of the common classifications and part of the student course manual of the internationally accepted ATLS trauma management system[@guly_vital_2011][@noauthor_atls_2018].

@@

The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24 hour of injury[@berry_shock_2015], as the midian time to death is 2 hours[@tisherman_detailed_2015]. To address these and other challenges within trauma care and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated. QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009]. By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally[@juillard_establishing_2009][@hashmi_hospital-based_2013].

@@

To classify, all of the parameters in one class needs to be fulfilled. In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017]. At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[@noauthor_atls_2018].
